---
output:
  bookdown::pdf_document2:
    latex_engine: xelatex
    toc: false
    number_sections: false
    fig_caption: yes
mainfont: Calibri
sansfont: Calibri
fontsize: 11pt
geometry: margin = 1in
linestretch: 2  
header-includes:
- \usepackage{setspace}\doublespacing
- \usepackage{dcolumn}
- \usepackage[labelfont=bf]{caption}
- \usepackage{float}
- \usepackage{afterpage}
- \usepackage{siunitx}
- \usepackage{amsmath}
- \usepackage{graphicx}
bibliography: references/GrasslandRefs.bib  # synced with zotero library
csl: references/journal-of-applied-ecology.csl # citation style
---

# Restoration of plant diversity in permanent grassland by seeding: assessing the limiting factors along land-use gradients

Martin Freitag^1^*, Valentin H. Klaus^2^, Ralph Bolliger^3^, Ute Hamer^1^, 
Till Kleinebecker^4^, Daniel Prati^3^, Deborah Schäfer^5^, Norbert Hölzel^1^

\ 

^1^Institute of Landscape Ecology, University of Münster, Heisenbergstr. 2, 48149 Münster, Germany

^2^Institute of Agricultural Sciences, ETH Zürich, Universitätstr. 2, 8092 Zürich, Switzerland

^3^Institute of Plant Sciences, University of Bern, 3013 Bern, Switzerland.

^4^Department of Landscape Ecology and Resource Management, 
Justus Liebig University Giessen, Heinrich-Buff-Ring 26-32, 35392 Gießen, Germany

^5^Botanical Garden of the University of Bern, Altenbergrain 21, 3013 Bern, Switzerland

\ 

*corresponding author: martin.freitag@uni-muenster.de

\ 

**Keywords**: dispersal limitation, fertilization, functional traits, land-use intensity,
productivity, seed addition, seedling establishment, trait-environment interactions


\newpage


## ABSTRACT
1.	Land-use intensification has caused dramatic declines in grassland biodiversity, 
and in Europe this decline has not yet been halted. 
Therefore, a major challenge in ecological restoration is 
to effectively restore plant diversity in permanent grasslands, 
i.e. long-term managed grassland not being used as cropland.
As species richness is often limited by seed dispersal,
introducing locally absent species is crucial. 
Seeding diverse seed mixtures is an appropriate tool, 
but restoration outcomes are highly variable.
Considering constraints by land-use intensity and plant functional traits 
during restoration could improve the efficacy of restoration efforts.

2.	We established a full-factorial experiment with high-diversity seeding and topsoil disturbance
in 73 grasslands along land-use intensity and productivity gradients 
and evaluated the restoration success over five years.
In addition, after five years we assessed the number of established species
and tested whether specific traits promoted the establishment of sown species in interaction with productivity.

3.	Plant diversity increased considerably when seeding and disturbance were combined,
while seeding alone only slightly increased species richness.
Along land-use gradients, we found no direct effects of grazing or mowing 
on the establishment of sown species, 
but fertilization indirectly limited establishment in the combined treatment
via a negative effect of productivity.

4.	Functional traits structured the establishment of sown species,
especially in the combined treatments, but effects varied with productivity. 
Species with low seed mass and low specific leaf area had greater establishment success
in low-productivity grasslands, whereas species with high seed mass
had greater establishment in high-productivity grasslands.


5. *Synthesis and applications*
Our results highlight the potential of using diverse seed mixtures for biodiversity restoration
in a wide range of permanent grasslands,
especially when combined with topsoil disturbance. 
However, productivity both limited restoration success
and affected establishment via an interaction with plant functional traits.
As productivity is a major constraint in grassland restoration,
we strongly recommend to first stop fertilization
and reduce soil nutrient loads in highly productive grasslands.
Restoration by seeding, with or without disturbance,
should only be applied when productivity is reduced
and the local environment should be considered
to compile site-specific seed mixtures to maximise restoration success.


\newpage


## INTRODUCTION 

Land-use intensification has dramatically impacted both biodiversity 
and ecosystem service provision of grasslands [@ipbes2019; @newbold2016].
Europe, for example, faces alarming and ongoing declines in area and deteriorations in conservation status
of permanent grassland [@janssen2016], 
i.e. continously managed grassland that was not used as cropland for at least five years.
Severe declines in plant species richness have been especially caused by intensive fertilization [@gross2009; @socher2012],
such that even common plant species have decreased in abundance at alarming rates [@jansen2019].
Grasslands therefore are a major target of ecological restoration [@blakesley2016].
Yet, our predictions of restoration success are often inaccurate
because outcomes vary greatly among restoration efforts, 
mostly because of widespread variability in restoration approaches and site conditions [@brudvig2017].

To restore the local plant species richness in grasslands, 
seed dispersal limitation must be overcome [@clark2007; @myers2009; @munzbergova2005].
Most grassland species are short-distance seed dispersers [@coulson2001; @sperry2019] 
and do not form persistent soil seed banks [@kiss2018].
Therefore, a mandatory step to restore plant species richness is to actively introduce locally absent species.
Seed sowing is a commonly applied technique that enables one 
to tailor the species composition to a target community [@kiehl2010].
In addition, diverse seeding could be implemented in agricultural routines
to increase diversity at larger scales.
Restoration by seeding has been successfully applied on ex-arable land 
[@pywell2002; @wagner2020] and in low- and 
high-productivity grasslands [@ladouceur2020; @pywell2007; @zobel2000].
Yet, most studies considered single or few sites, 
which makes it difficult to assess the significance of key constraining factors
across real world gradients of land-use intensity and productivity.

In the germination and seedling stages of a plant, 
safe microsites such as gaps within the established vegetation are crucial [@grubb1977].
These gaps in the established sward are often created via topsoil disturbance, 
which enables seed-soil contact and amplifies seed germination 
through reduced competition and increased light availability
at the soil surface [@bischoff2018; @myers2009].
Such microsites create a safe regeneration space even in productive grasslands [@foster2004],
but competition with the regrowing vegetation sward in the years after disturbance
may impede long-term establishment in productive sites [@dickson2008].
This competition for light can be reduced with repeated biomass removal
through mowing or grazing, thereby increasing the establishment success [@dickson2008].
Additionally, grazing creates microsites by trampling, 
which could further facilitate establishment [@kladivova2016].
Although high-diversity seed mixtures increased diversity even in the long run [e.g. @maccherini2012], 
it remains largely unclear how topsoil disturbance interacts with grazing or mowing 
to shape establishment in managed grasslands.
Determining how topsoil disturbance, grazing and mowing 
affect the long-term establishment of seeds along productivity gradients 
can improve outcome predictions and may provide guidance 
on how to prioritize sites for restoration actions.

To improve restoration success and maximise ecosystem functions, 
plant functional traits can help to compile targeted seed mixtures [@laughlin2014]. 
Because traits, especially canopy height, specific leaf area and seed mass,
represent fundamental trade-offs of plant strategies in response to the environment
and affect competitive and reproductive abilities [@westoby1998],
they can be used, for example, to select potentially successful species to increase restoration success [@pywell2003].
Further, a deeper understanding of assembly mechanisms in restoration 
enables us to restore communities targeted to specific environmental conditions [@laughlin2014; @torok2017].
Assembly mechanisms such as environmental filtering influence restoration success 
because it structures the establishment of species with particular traits [@balazs2020; @zirbel2020].
Knowledge on how traits interact with environmental gradients, 
such as productivity and land-use intensity, 
can assist in the selection of species based on local site conditions.

Here, we aimed to assess the long-term restoration success using a diverse 
and regionally adapted seed mixture along gradients of land-use intensity and productivity, 
with and without prior topsoil disturbance.
We established a full-factorial experiment in 73 grasslands 
ranging from low-productivity sheep pastures and single-cut meadows 
to highly fertilized meadows and mown pastures with four cuts pear year or intensive grazing.
Short-term results from this experiment showed that intensively used, 
species-poor grasslands increased strongly in species richness 
when seeding and topsoil disturbance were combined, 
while seeding alone had no effect on diversity [@klaus2017].
Here, we assessed how land use, productivity and plant functional traits 
structure the long-term establishment of sown species.
Specifically, we investigate i) the effectiveness of seeding, topsoil disturbance
and their combination to enhance plant diversity over time,
ii) the degree to which land-use intensity and productivity 
constrain the establishment of sown species 
and iii) how traits, trait-environment and trait-management interactions 
affect the establishment of sown species during restoration.

\  

## METHODS

### Study design

We analysed five years of a seeding and topsoil disturbance experiment in 73 grasslands
as part of the Biodiversity Exploratories project [@fischer2010].
We installed the experiment in three German regions that span gradients in soil characteristics,
elevation and climate that are representative of large parts of Central Europe. 
The regions comprise 
i) the calcareous mid-mountain range Biosphere Reserve Schwäbische Alb (48.4°N, 9.4°E), 
ii) the calcareous low-mountain range National Park Hainich-Dün and surroundings (51.1°N, 10.4°E) 
and iii) the postglacial landscape of Biosphere Reserve Schorfheide-Chorin (53.0°N, 14.0°E).
While grasslands in the Schwäbische Alb and Hainich-Dün are restricted to mineral soils,
grasslands in Schorfheide-Chorin are situated on drained fen soils
or slightly acidic sandy soils [@fischer2010].
Within each region, we selected 23-25 permanent grasslands along gradients of land-use intensity, 
productivity and corresponding plant diversity, ranging from unfertilized sheep pastures 
with 70 species per 16 m^2^ to highly fertilized, productive 
and frequently mown grasslands with 10 species per 16 m^2^ [@socher2012].
To quantify land-use intensity, farmers were interviewed each year 
to inquire on the amount of fertilizer used (in kg N ha^-1^), 
mowing frequency per year and grazing intensity (livestock unit grazing days × ha^-1^) [@vogt2019].
Here, we used averaged fertilization, mowing and grazing intensities from 2014 to 2018.

We established a 2 × 2 factorial experiment with seeding and topsoil disturbance in 73 grasslands.
This resulted in four 7 m × 7 m treatments (control, seeding only, 
seeding and disturbance, disturbance only), 
which we arranged in a rectangle with 2 m distance between treatments (see appendix S1). 
In October 2014, we applied the disturbance treatment by mechanical perturbation of the established sward 
and the topsoil down to 10 cm with a rotary harrow or a rotary cultivator.
The topsoil disturbance created a high proportion of bare ground 
[~50% in the following spring; @schafer2019] 
and was similar to common agricultural re-seeding practices.
We did not remove root and shoot fragments of the disturbed sward so that plants could possibly regrow.
See @klaus2017 for further details on the experimental design.

For the seeding treatments, we used commercially available regional seed mixtures 
from certified seed producers, 
composed of genotypes that are assumed to be adapted to regional environmental conditions [@bucharova2017].
The mixtures consisted of common and less-common species,
selected from the species pools of each individual region  [@fischer2010]
and included grasses, legumes and forbs.
Seeding density per species depended inversely on seed mass 
(high: 100 seeds m^−2^ if seed mass < 1 mg, 
medium: 34 seeds m^−2^ between 1 mg and 10 mg, low: 17 seeds m^−2^ > 10 mg). 
Two-thirds of the mixture were sown in November 2014 and one-third in March 2015. 
In total, we sowed 5.37 g m^-2^ and 66 species in Schwäbische Alb, 
4.11 g m^-2^ and 52 species in Hainich-Dün and 3.47 g m^-2^ 
and 47 species in Schorfheide-Chorin (no regional seeds of legumes available 
from seed producers for this region).
To estimate germination rates, we sowed 60 seeds per species and region
in standard potting soil.
We placed the pots outdoors from December 2015 until March 2014 in Bern, Switzerland, 
to allow for cold stratification, relocated the pots thereafter to a greenhouse
and counted the emerging seedlings over three months.


We surveyed vegetation and measured productivity from 2015 to 2019 annually in May 
before the first mowing or grazing took place.
On all four 7 m × 7 m treatments in each of the 73 grasslands, 
we recorded all vascular plant species on 2 m × 2 m quadrats and estimated percentage cover.
As a measure of competition,
annually we clipped all living plant biomass on one square meter on the control treatments
(four 0.25 m^2^ squares, different but close-by locations each year, see appendix S1).
We dried the biomass for 48 h at 80 °C and weighed it to the nearest gram. 
In 2018, we could not access three grasslands due to early land use 
(i.e. missing data for 12 observations).
In total, we recorded 1448 observations over the five years.


### Analysis

*Seeding and topsoil disturbance effects on plant diversity*

We analysed how seeding and topsoil disturbance influenced plant diversity over time. 
We used two complementary diversity indices, namely species richness 
and effective number of species S~PIE~, which accounts for community evenness 
by weighing abundant species more than rare species.
We calculated evenness as the probability of interspecific encounter $PIE = \sum_{i=1}^{S} p_i^2$,
where $S$ is species richness and $p_i$ is the relative cover of species $i$,
and transformed $PIE$ into an effective number of species 
$S_{PIE} = \frac{1}{1 - PIE}$ [@chase2018; @jost2006]. 
S~PIE~ equals species richness when all species have equal abundance, 
but S~PIE~ decreases when cover is more unevenly distributed across species.
We modelled species richness and S~PIE~ as a function 
of seeding, disturbance and year, including all their interactions,
and assumed Poisson error structure for species richness and 
Gamma error structure for S~PIE~, both with log link.
We treated year as categorical because effects over time were non-linear.
We included varying intercepts for grassland site to account for the experimental design 
and varying intercepts for year within region (five years times three regions) 
because weather conditions and observers varied between years and regions.
The multi-level models were estimated in a hierarchical Bayesian framework 
using the Stan probabilistic language [@standevelopmentteam2020] 
accessed via the package *brms* v2.13.0 [@burkner2017] in R v4.0.3 [@rcoreteam2020].
For all models, we specified weakly informative normal priors with zero mean for fixed parameters 
and default vague priors for the intercepts (see appendix S2).
We ran four parallel chains and 10,000 iterations (5,000 discarded as burn-in) 
and ensured convergence with R-hat values being <1.01 for all parameters.
Model fit was assessed with posterior-predictive checks 
using the package *bayesplot* v1.7.2 [@gabry2019; @gabry2019a; see appendix Fig. S2].
We report 90% credible intervals (CrI) for posterior samples, 
which do not overlap zero when 95% of the posterior is either above or below zero, 
and we calculated conditional and marginal Bayes $R^2$ [@gelman2019] 
to assess how much variation is explained by the treatments or varying intercepts.


*Direct and indirect effects of land use on establishment*

To evaluate how grazing, mowing and productivity modify or create 
regeneration spaces and hence determine the number of established seeding species, 
we modelled the absolute difference in sown species richness 
between the seeding treatments and control ($\Delta$richness)
as well the difference in total species richness.
We fitted two separate multivariate models for seeding (only) 
and seeding and topsoil disturbance (combined) treatments
in the first and in the fifth year.
Multivariate models sample the parameters jointly for the sub-models and, thus, allow modelling 
response variables measured on the same experimental unit.
We expected direct effects of grazing (log-transformed grazing days ha^-1^), 
mowing frequency and productivity (log-transformed g m^-2^) on establishment and 
included region effects as well as an effect of $\Delta$richness 
of year one on year five (see also Fig. 2).
Because fertilization intensity (log-transformed kg Nitrogen ha^-1^) 
was closely correlated to mowing frequency [$\rho$=0.66, @vogt2019], 
we included only an indirect effect of fertilization on $\Delta$richness via productivity.
We modelled productivity with Gamma distribution and log link 
and assumed Gaussian errors for $\Delta$richness.
The repeated biomass measurements were simultaneously modelled 
as arising from a normal distribution with a mean equal to the 
‘true’ latent productivity of each grassland site and measurement error $\sigma$.
We chose arithmetic means and standard deviations of harvested biomass 
as informed priors for the means of latent productivity, (i.e., varying between grasslands) 
and a weakly informative prior for the measurement error $\sigma$ 
(see appendix S4).
The multivariate models were estimated using *rstan* v2.19.3 [@standevelopmentteam2020]. 


*Plant functional traits effects on establishment*

To explore how functional traits affect the establishment of sown species 
five years after seeding along the productivity gradient, 
we modelled establishment with logistic regression.
We assumed sown species to have successfully established 
in the seeding treatments if they were present in year five.
Because some of the sown species may have been present before the start of the experiment, 
we only judged whether a species successfully established when the species was absent 
on either the control or disturbance-only treatment in any year.
We kept 2,746 observations, because sown species were already present
in the respective grasslands in 37% of the cases in the Alb region,
36% in the Hainich region and 21% in the Schorfheide region (see Tables S3 and S4).
This definition is conservative, especially for common species, 
but reduces the number of false positive observations where a sown species was already present. 
From the LEDA Traitbase [@kleyer2008] we extracted specific leaf area (relative growth rate), 
canopy height (competitive ability of the adult plant) 
and seed mass (energy for seedling recruitment) and averaged the trait values by species.

To test for the effects of traits and trait-environment interactions on establishment, 
we modelled binary establishment with logit link function as a function of 
height (log.), seed mass (log.) and specific leaf area. 
We included estimated live seeding density (seeding density multiplied by germination rate)
as a continuous predictor.
Informed by a negative effect of productivity on the number of established species, 
we initially included productivity (log-transformed, as a measure of competition) 
and interactions of traits with productivity.
We included varying intercepts for both species identity and grassland site 
due to the multi-level structure and scaled all predictors to unit standard deviation. 
Productivity was again modelled as latent 
(see appendix S5). 

To test for trait-grazing and trait-mowing effects,
we fitted three additional models with i) trait-grazing interactions, 
ii) trait-mowing interactions and iii) both trait-grazing and trait-mowing interactions
in addition to trait-productivity interactions. 
To judge whether including grazing and mowing interactions improved
the prediction of establishment, we compared models based on 
the expected log point-wise predictive density from leave-one-out cross-validation [@vehtari2017; @vehtari2020].

\  

## RESULTS

*Seeding and topsoil disturbance effects on plant diversity*

High-diversity seeding had clear positive effects on species richness
over the five years of this study, but less so on S~PIE~,
which accounts for community evenness. 
The seeding effect on species richness increased steadily over time,
from slightly positive in the first year to clearly positive in the fifth year 
(seeding $\beta$=0.04, 90% CrI: -0.0 to 0.1, 
plus seeding\*5th year $\beta$=0.09, 90% CrI: 0.0 to 0.2; Fig. 1a; appendix Table S1). 
Transformed to the response scale with average species richness of 25.7 per 4 m^2^ on the control, 
this resulted in an increase in 3.1 species in year five (Fig. 1c).
Seeding alone had no clear effect on S~PIE~ (Fig. 1a, f), 
which suggests that increases in richness were compensated by decreases in evenness
because the sown species established with low abundance.
Topsoil disturbance led to temporary increases in both species richness and S~PIE~ (Fig. 1a, b, e),
indicating the germination of ruderal species and an increase in evenness after disturbance.

Both species richness and S~PIE~ considerably increased when seeding and topsoil disturbance were combined.
For species richness, the seeding and disturbance interaction was clearly positive in the first year
but became weaker over time (seeding\*disturbance $\beta$=0.25, 90% CrI: 0.2 to 0.3, 
plus seeding\*disturbance\*5th year $\beta$=-0.16, 90% CrI: -0.3 to -0.1; Fig. 1a, d; appendix Table S1). 
S~PIE~ was also positively affected by the seeding\*disturbance interaction 
(seeding\*disturbance $\beta$=0.15, 90% CrI: 0.0 to 0.3), 
but temporary increases in S~PIE~ were mainly driven by the disturbance effect
(e.g. disturbance\*2nd year $\beta$=0.15, 90% CrI: 0.0 to 0.3; Fig. 1a, e). 
On the response scale, the combination of seeding and disturbance led 
to an increase in species richness by 8.3 species in the fifth year (Fig. 1d) 
and an increase in S~PIE~ by 0.6 (Fig. 1g), given an average S~PIE~ of 6.2 on the control.

*Direct and indirect effects of land use on establishment*

We found no direct effects of grazing or mowing on the number of established species ($\Delta$richness),
but fertilization indirectly limited the long-term establishment on the combined treatment 
via a negative effect of productivity on $\Delta$richness in year five
(Fig. 2b, d; appendix Tables S2 and S3).
$Delta$richness in the first year was less related to productivity or land use, 
underlined by lower $R^2$ values for both treatments compared to the fifth year (Fig. 2).
Instead, $\Delta$richness of the combined treatment in the first year 
was higher when considering all species (Intercept $\alpha$=9.84, 90% CrI: 8.4 to 11.2; Fig. 2b) 
compared to $\Delta$richness of seeding species (Intercept $\alpha$=6.10, 90% CrI: 5.2 to 7.0; Fig. 2d),
which suggests a considerable but temporary contribution of ruderal species 
to $\Delta$richness after topsoil disturbance.
However, sown species $\Delta$richness increased until the fifth year,
and the difference to $\Delta$richness considering all species decreased considerably (Fig. 2b, d). 
The negative effect of productivity on $\Delta$richness was less clear for the seeding-only treatment,
but turned from neutral to weakly negative over time for sown species $\Delta$richness (Fig. 2a, c).

*Plant functional traits effects on establishment*

Plant functional traits considerably influenced establishment, 
but effects of traits on establishment interacted with productivity. 
Establishment was consistently predicted by productivity, 
specific leaf area and their interaction  (Fig. 4c, d).
Productivity decreased establishment drastically 
such that the probability of successful establishment 
in year five was almost zero in productive grasslands for the
seeding-only and seeding and topsoil disturbance treatments (insets in Fig. 4). 
Species with low specific leaf area had higher average establishment success,
but trait-productivity interactions for both seeding treatments 
indicate that species with low specific leaf area 
established increasingly better at low-productivity sites (Fig. 4c, f).
The effects of height and seed mass were less unclear (Fig. 4a, b, d, e), 
but the trait-productivity interactions suggest that tall species with high seed mass
established better at highly productive sites in the combined treatment (Fig. 4d, e).
We tested for additional trait-grazing and trait-mowing interaction effects 
on establishment, but based on leave-one out cross-validation, 
neither interaction improved the prediction of establishment success (appendix Table S7). 

Live seeding density (mean 7.5 viable seeds m^-2^, appendix Fig. S6) increased establishment rates
for both seeding-only ($\beta$=0.68, 90% CrI: 0.4 to 0.9, appendix Table S4) 
and seeding and topsoil disturbance treatments ($\beta$=0.88, 90% CrI: 0.7 to 1.1).
The effects on establishment on the seeding-only treatment were less clear, as
overall parameter uncertainty was higher and average establishment success lower 
(Intercept $\alpha$=-4.02, 90% CrI: -4.7 to -3.4) 
compared to the combined treatment (Intercept $\alpha$=-2.57, 90% CrI: -3.1 to -2.1).
The marginal Bayes $R^2$ was considerably lower than the conditional $R^2$ for both seeding treatments
(seeding only: median $R_{marg}^2$=0.20, $R_{cond}^2$=0.47; 
seeding and disturbance: median $R_{marg}^2$=0.29, $R_{cond}^2$=0.46), 
which implies that grassland site and species identity 
accounted for a considerable share of variance not explained by traits. 


\  

## DISCUSSION

The ongoing biodiversity crisis reflects an urgent need 
to restore plant diversity in grasslands wherever possible, 
but predicting restoration success is remarkably difficult [@brudvig2017],
and restoration actions need to be carefully adapted to local site conditions [@laughlin2014; @torok2017].
Our results highlight the potential for using high-diversity seed mixtures 
to restore plant diversity across a wide range of permanent grasslands, 
especially when combined with topsoil disturbance.
However, productivity limited the overall establishment success of sown species 
and shaped establishment via an interaction with plant functional traits. 
We demonstrate that productivity is a major constraint 
for the restoration of plant diversity in managed grasslands,
but productivity can be used as an indicator to plan restoration actions 
and select suitable target species along productivity gradients.

The most effective method for increasing plant diversity
was the combination of seeding and topsoil disturbance. 
This finding is well in line with early results from this experiment [@klaus2017] 
and confirms previous studies, which conclude that species richness 
is largely limited by dispersal limitation and the lack of a persistent soil seed bank
for most species [@clark2007; @klaus2018; @myers2009].
Concurrent with previous research, disturbing the topsoil reduced microsite limitations
and effectively created regeneration space for the seeded species to establish [@bischoff2018; @grubb1977; @munzbergova2005].
Yet, over the five years of this study, species richness slowly increased 
in the seeding-only treatment as well.
This suggests that at least some species can establish without topsoil disturbance, 
for instance when microsites are created by grazing [@kladivova2016; @maccherini2012].

While we observed clear seeding effects on species richness, 
we found weaker effects on the effective number of species S~PIE~,
indicating that increases in richness were partly outweighed by decreases in community evenness 
due to low cover of the newly established species.
In contrast to @ladouceur2020, who reported positive effects of seeding-only on S~PIE~
across a range of natural and semi-natural grasslands, 
in our study the lack of a seeding-only effect on S~PIE~ suggests that 
when topsoil disturbance was not applied, sown species 
had not yet become important members of the community. 
A possible explanation is that a higher average productivity of our grasslands
limits establishment more strongly compared to the partially degraded grasslands 
included in the study by @ladouceur2020.
The weaker positive effects of the seeding treatments on S~PIE~ due to increases in evenness
suggest that creating microsites is important for the introduced species to establish 
with noticeable cover, but it remains an open question as to how long these differences persist.

We found no direct effects of land-use intensity on the number of established seeding species, 
but fertilization indirectly limited long-term establishment 
via a negative effect of productivity. 
At first, the intensively used productive grasslands 
benefited most from seeding and topsoil disturbance [@klaus2017], 
but initial increases in species richness were recorded right after the set-up of the experiment,
before the first mowing or grazing management,
and were partly driven by ruderal species from the soil seed bank. 
While previous studies reported that productivity 
limits seeding successes in grasslands [@dickson2008; @foster2001],
our study is the first to confirm this pattern along a realistic grassland productivity gradient. 
The increasing impact of productivity on establishment over time
suggests that enhancing plant diversity through seeding 
is only an ephemeral success in productive grasslands 
and highlights that recruitment limitation is an at least as important constrain
for species richness as dispersal limitation  [@clark2007; @munzbergova2005].
Concurrently, the number of established species 
correlated positively with the resident species richness (see appendix S6), 
both of which are constrained by productivity [@socher2012].
Thus, productivity limits both resident species richness and the establishment of introduced species,
indicating that both species richness and the potential 
to introduce new species decrease with higher fertilization, which increases productivity.

Several studies have emphasized that soil fertility 
and its determinant role for productivity as one of the most important constraints
to restoring species-rich grasslands [@dickson2008; @walker2004],
and fertilization significantly contributes to this constraint. 
However, we were not able to identify a threshold of productivity 
at which restoration would change from successful to failed.
Our results suggest that the establishment success is generally low 
at productivity levels above ~250 g m^-2^ aboveground green biomass in May, 
but the establishment probably still depends 
to a significant extent on further site characteristics. 
One such factor that determines seedling establishment might be litter biomass. 
While small amounts of litter can facilitate seedling recruitment in dry grasslands, 
dense litter (>500 g m^-2^) largely inhibits recruitment [@loydi2013], 
However, we rarely observed significant amounts of litter in our grasslands,
which is typical for regularly used Central European grasslands [@gross2009].

Interestingly, grazing and mowing had no effect on the number of established species.
Grazing is assumed to create microsites by feeding and trampling and, hence, 
can increase the recruitment from seeds [@kladivova2016],
but effects of grazing on establishment are usually low [@maccherini2012; @rasran2007].
In our study, bare soil cover was hardly related to grazing or mowing intensities (appendix Fig. S9).
We suspect that productivity partly reflects the availability of microsites,
because sparse vegetation with low aboveground biomass provides more space per se, 
or because the effects of grazing and mowing on microsites
were ovrruled by productivity.
Based on our results, we conclude that, in fertile grasslands, 
productivity must be reduced to ensure restoration success.
The main focus in productive grasslands, therefore, must be 
to stop fertilization and reduce the soil nutrient loads,
before fine-tuning the grazing and mowing management.

Plant functional traits clearly affected establishment.
Sown species with a low specific leaf area consistently 
established better, independent of productivity.
However, the patterns we observed for the seeding-only treatment were less clear, 
probably due to generally lower establishment rates 
and subsequently greater uncertainty in the relationships between traits and establishment.
When seeding was combined with topsoil disturbance, we also found that low-competitive species 
with low canopy height and low seed mass 
(such as *Dianthus carthusianorum* or *Sanguisorba minor*) tended to establish better.
In part, these main effects were possibly driven by the fact that 
the overall establishment rate decreased with productivity. 
At low levels of productivity, the plant community is characterized by 
species with a low specific leaf area [@allan2015],
as conservative and slow-growing species with low specific leaf area 
have no disadvantage in establishing and are possibly better adapted to less fertile environments.

Such adaptions to the environment are reflected in the interactions 
between functional traits and productivity that shape establishment during restoration. 
For instance, in the combined treatment, species with high seed mass 
established better in productive sites. 
Having considerable energy stored in large seeds increases recruitment rates [@clark2007] 
and seedling survival rates [@metz2010], 
which seems to be a particular advantage in productive sites,
where competition with the regrowing vegetation selects for the most competitive seedlings. 
In contrast, species with low specific leaf area 
established better at sites of low productivity.
The fast resource-acquisition strategy of species with high specific leaf area
is beneficial under nutrient-rich conditions, for instance resulting from intensive fertilization [@allan2015;@vries2012].
Our results confirm the finding that this adaption to nutrient surplus in productive sites
does not represent an advantage under conditions with less competition for light [@zirbel2020],
where, for instance, water availability limits plant growth.
We propose that sown species in grasslands are environmentally filtered by productivity 
and successful establishers possibly reflect the traits of the resident plant community [@breitschwerdt2015;@ames2020].
Variation in establishment success along environmental gradients 
and trait-environment interactions are rarely studied, 
but insights on these relationships provide valuable information
to prioritize restoration sites and compile site-specific seed mixtures
to maximise success and ecosystem functions during restoration [@laughlin2014; @balazs2020]

\  

## ACKNOWLEDGEMENTS

We thank Judith Hinderling, Torsten Meene, Svenja Kunze and many students for help during fieldwork
and three anonymous reviewers for very constructive comments.
We also thank the managers of the three Exploratories, Kirsten Reichel-Jung, 
Iris Steitz, Sandra Weithmann, Florian Straub, Katrin Lorenzen, Juliane Vogt, 
Martin Gorke and Miriam Teuscher for their work in realizing this experiment 
and maintaining the plot and project infrastructure, Christiane Fischer 
and Jule Mangels for giving support through the central office, Michael Owonibi 
and Andreas Ostrowski for managing the central database, and Eduard Linsenmair, 
Dominik Hessenmöller, Jens Nieschulze, Francois Buscot, Ernst-Detlef Schulze, 
Wolfgang W. Weisser, and the late Elisabeth Kalko for their role 
in setting up the Biodiversity Exploratories project.
The work has been funded by the German Research Foundation (DFG) Priority Program 1374 
‘Infrastructure-Biodiversity-Exploratories’ (FI 1246/15-2; HA 4597/6-3; HO 3830/2-4; KL 2265/4-4). 
Field work permits were issued by the responsible state environmental offices 
of Baden-Württemberg, Thüringen, and Brandenburg (according to § 72 BbgNatSchG).

\  

## AUTHORS’ CONTRIBUTIONS

D.P., U.H., N.H., V.K., T.K. and M.F. conceived the ideas and designed the experiment, 
V.K., D.S., R.B. and M.F. collected the data, M.F. analysed the data 
and led the writing of the manuscript.
All authors contributed critically to the drafts and gave final approval for publication.

\  

## DATA AVAILABILITY STATEMENT

All data and code used for this study will be archived within the BExIS database 
of the Biodiversity Exploratories project (https://bexis.uni-jena.de/PublicData).

\newpage

## REFERENCES

<div id="refs"></div>  <!-- put the references here --> 

\newpage


## FIGURES AND TABLES



```{r treatments, out.width = "100%", fig.cap = captreatments, echo=FALSE}
  knitr::include_graphics("figs/01-1_treatment-effects.pdf")
captreatments <- "
Effects of seeding and topsoil disturbance 
on species richness and S~PIE~ over time. 
(a) Effect sizes of seeding, disturbance and the seeding and disturbance interaction, 
as the main and interacting effects with years, for species richness and S~PIE~ 
(intercept, year main effects and varying intercepts not shown, see appendix Table S1). 
Both diversity measures were modelled with log-link assuming
Poisson and Gamma error structure, respectively.
Panels (b) to (g) display differences between treatments and 
the control on the response scale. Seeding alone had small positive effects on species richness, 
which increased over time (c), while this was not found for S~PIE~ (f).
The combination of seeding and disturbance had a clear positive effect 
on both diversity metrics, although it was weaker for S~PIE~.
To calculate differences between treatments and control, 
we predicted species richness and S~PIE~ based on additive parameter effects (a), 
transformed predicted values to the response scale
and calculated the difference in predicted values between treatments and control.
Points indicate medians along with 90\\% CrI."
```


\newpage


```{r sem, out.width = "100%", fig.cap = capsem, echo=FALSE}
  knitr::include_graphics("figs/01-2_land-use-delta-richness.pdf")
capsem <- "
Direct and indirect land-use effects on the number of established species
($\\Delta$richness, the difference in species richness between treatment and control)
for total species richness and sown species richness.
Land-use effects on $\\Delta$richness were absent on the seeding-only treatment (a, c).
Fertilization had an indirect negative effect via productivity on $\\Delta$richness in the fifth year
for both (b) total and (d) sown species richness on the seeding and disturbance treatment. 
In the first year after disturbance, the larger $\\Delta$richness for total species richness 
indicates a considerable increase in unsown species on the seeding and disturbance treatment, 
which almost completely vanished in the fifth year.
Positive paths are shown as solid and negative paths as dashed lines 
along with standardized regression coefficients and 90\\% CrI, 
in blue (positive) and red (negative) when the CrI does not overlap zero.
Fixed effects of region are displayed in the $\\Delta$richness boxes."
```

\newpage


```{r productivity-predictions, out.width = "60%", fig.cap = capproductivitypredictions, echo=FALSE}
  knitr::include_graphics("figs/01-3_delta-sown-richness-predict.pdf")
capproductivitypredictions <- "
Conditional effect of productivity on 
the number of established sown species as compared
to the control ($\\Delta$richness) in the fifth year.
(a) Overall sown species $\\Delta$richness was lower 
and the effect of productivity weak on the seeding-only treatment, 
(b) but sown species $\\Delta$richness declined with productivity on the combined treatment. 
Lines denote the median and 90\\% CrI of sown species $\\Delta$richness 
across all 73 grasslands (grey) and within regions (coloured), 
conditioned on mean values of the other predictors (see Fig. 2).
"
```


\newpage


```{r trait-predictions, out.width = "100%", fig.cap = captraitpredictions, echo=FALSE}
  knitr::include_graphics("figs/01-4_traits-predict.pdf")
captraitpredictions <- "
Conditional effects of plant height, 
seed mass, specific leaf area and productivity (log scale) on the establishment 
of sown species on (a)-(c) the seeding-only and (d)-(f) the seeding and topsoil disturbance treatments. 
Especially in the combined treatment, trait effects on establishment 
interacted considerably with productivity. 
Coloured lines denote the median and 90\\% CrI for low (mean minus one SD, red) 
and high (mean plus one SD, blue) trait values, 
conditioned on mean values of the other traits and mean sowing density.
In each panel, inset plots show the median and and
90\\% CrI of the parameters used for prediction.
See Table S4 for details."
```




\clearpage

